Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-01-19 jss/EpiCompare-paper/EpiCompare-paper.tex
Update to section 3.

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{xcolor}

@@

\section[Package overview]{Package tool overview}\label{sec:overview}

@@

In this section we will present the tools in this package and how they
aid in the data analysis pipeline. We present a modification of Figure
\ref{fig:pipeline} from the Section \ref{sec:intro} with Figure
\ref{fig:pipeline2} with uses of our package. We recommend the package
user keep this image around as a reference. All user functions are aimed
to be as friendly as possible, and we focus on providing the user
``tidyverse'' style functions, that encourage piping and also follow
clear verb naming schemes \citep{Wickham2019}. There are 2 different
ways \pkg{EpiCompare} can be incorporated in the data analysis pipeline
for epidemics, either at the very beginning when pre-processing data and
visualizing raw data, or after modeling has been done and a generative
model is fit. Figure \ref{fig:pipeline2} captures these different paths,
and we will highlight both approaches and how to leverage
\pkg{EpiCompare} in the subsections below.

\begin{figure}[!ht]
    \centering
    \includegraphics[width = 1\textwidth]{images/pipeline2.png}
    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
    \label{fig:pipeline2}
\end{figure}

\hypertarget{data-preprocessing}{%
\subsection{Data Preprocessing}\label{data-preprocessing}}

The first step of most data analysis is cleaning up the data to be
explored. There are multiple ways to collect the epidemiological data.
Sometimes individual records with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and sub-population information will be
collected, whereas other data collections will focus aggregate counts of
individuals in each epidemic state. In fact, usually only the number of
new infections at each time step is observed and the compartment totals
are then imputed from that along with other information about the
disease and the population of interest. We focus on understanding the
overall impact of an outbreak at the aggregate/population level. We
provide a function to transform information about each
individual/agent's start time of each state (e.g.~start of infection,
etc). This transformation between agent information to aggregate
information is useful for seeing the overall trend of the epidemic (and
how it impacts different subpopulations). Our tools aim to allow the
user to easily group agents and define new subpopulations to explore.
This is important as case studies highlight the usefulness of
identifying differing subpopulations (e.g.~age, sex) and many state
based models provide for subpopulation-based states in their analysis
\citep{rvachev1985,anderson1992,worby2015}.

We provide a ``tidyverse''-styled function,
\texttt{agents\_to\_aggregate} to transform agent information into
aggregate state information, As a ``tidy'' function, our function
\texttt{agents\_to\_aggregate} allows the user to identify any
subpopulations by first doing \texttt{group\_by} from \texttt{dplyr},
and also allows for infinite epidemic states - which can be for models
with assumptions of sub population based states (e.g.~youth infections),
as well as indicators for death and birth dates. Our function,
\texttt{agents\_to\_aggregate} allows for states to be skipped -
captured with \texttt{NA} values and multiple starts of a state -
(e.g.~multiple infections) as long as the original data is stored in
columns that are constrained to ordering
ideas\footnote{make this more clear?}. Currently this function is
constrained to integer time steps (for example days), but
transformations (linear or otherwise) of the time columns can make the
time steps more (or less) . \texttt{agents\_to\_aggregate} then returns
a standardized count for each state for each time point. We see
\texttt{agents\_to\_aggregate} as a really powerful tool, but also very
useful to quickly transform agents information to be used in aggregate
state-based models and compare state-based models to observed data.

\hypertarget{eda}{%
\subsection{EDA}\label{eda}}

With raw data, getting to know our data now-a-days very frequently means
figuring out good combinations of visualizations, numerical summaries
and subsetting. An expert coder can start from
\texttt{agents\_to\_aggregate} to successfully do this in many ways, but
we've also developed tools to rapidly explore your data if your have a 3
unique states models (like the SIR model). Our \texttt{geom\_aggregate}
provides a rapid way to explore different subpopulations experience of
the epidemic, and combines the ideas behind agents\_to\_aggregate for
the SIR case, with \texttt{geom\_path} and \texttt{coord\_term} to
visualize any number of groups epidemic trajectory in 3d simplex space
using \texttt{ggplot2} and \texttt{ggtern}
\citep{Wickham2016, Hamilton2018}. Visualization tools for SIR models
were developed because (1) SIR models are some of the most common and
basic epidemic state-based models and (2) our simplex representation of
these epidemics emphasizes a ``time-invarance'' representation of the
data (for a refresher see Section \ref{sec:time-invariant}).

\hypertarget{model-fitting-and-simulations}{%
\subsection{Model Fitting and
Simulations}\label{model-fitting-and-simulations}}

Although this package does not focus on estimating a model for the data,
we do provide some power functions for simulation of basic state models
with Bernoulli approximation of ODE state models encoded in
simulate\_agents and simulate\_SIR\_agents (which can naturally be
combined with agents\_to\_aggregate ). \textbf{Description of these
models, reference to Shannon R0 paper, etc\ldots{}}

\hypertarget{post-processing}{%
\subsection{Post-processing}\label{post-processing}}

\ldots{}

\hypertarget{comparions-and-assessment}{%
\subsection{Comparions and Assessment}\label{comparions-and-assessment}}

\ldots{}

>>>>

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{xcolor}
\usepackage{rotating}

@@

\section[Package overview]{Overview of
\pkg{EpiCompare}}\label{sec:overview}

@@

In this section, we present the tools implemented in \pkg{EpiCompare}
and explain how they aid in the data analysis pipeline. In Fig.
\ref{fig:pipeline2}, we illustrate how our package fits into the data
analysis pipeline introduced in Fig. \ref{fig:pipeline}. All
front-facing functions are aimed to be as user-friendly as possible, and
we focus on providing the user ``tidyverse'' style functions, that
encourage piping and also follow clear verb naming schemes
\citep{Wickham2019}. Although users can typically incorporate
\pkg{EpiCompare} into any step in the data analysis process, there are
two primary points of entry. The first point of entry is the very
beginning with pre-processing and visualizing raw data, and the second
point of entry is after modeling and simulation. Figure
\ref{fig:pipeline2} captures these different paths, and we will
highlight both approaches and how to leverage \pkg{EpiCompare} in the
subsections below.

\begin{sidewaysfigure}[!ht]
    \centering
    \includegraphics[width = 1\textwidth]{images/pipeline2_v2.pdf}
    \caption{How \pkg{EpiCompare} supplements and aids in the epidemiological data analysis pipeline.}
    \label{fig:pipeline2}
\end{sidewaysfigure}

\textbf{Data Pre-processing}

The first step of most data analysis is cleaning up the data to be
explored. There are multiple ways to collect the epidemiological data.
Sometimes individual records with times of different states of the
epidemic (infection, recovery, etc.) as well as individual information
like network structure, location, and sub-population information will be
collected, whereas other data collections will focus on aggregate counts
of individuals in each epidemic state. In fact, usually only the number
of new infections at each time step is observed and the compartment
totals are then imputed from that along with other information about the
disease and the population of interest. We focus on understanding the
overall impact of an outbreak at the aggregate/population level. In
\pkg{EpiCompare}, we provide a function to transform information about
each individual/agent's initial time of entry into each state
(e.g.~start of infection, start of recovery, etc). This transformation
between \textit{agent} information to \textit{aggregate} information is
useful for seeing the overall trend of the epidemic (and how it impacts
different subpopulations). Our tools allow the user to easily group
agents and define new subpopulations to explore. This is important as
case studies highlight the usefulness of identifying differing
subpopulations (e.g.~age, sex) and many state based models provide for
subpopulation-based states in their analysis
\citep{rvachev1985,anderson1992,worby2015}.

We provide a ``tidyverse''-styled function,
\texttt{agents\_to\_aggregate()} to transform agent information into
aggregate state information. As a ``tidy'' function, our function
\texttt{agents\_to\_aggregate()} allows the user to identify and analyze
subpopulations by first executing a \texttt{group\_by()} from
\pkg{dplyr}. The function \texttt{agents\_to\_aggregate} is flexible in
that there is no limit on the amount of permitted epidemiological states
so users can incorporate information from compartments such as
``Exposed'', ``iMmune'', or ``Hospitalized'' groups, for instance, and
also permits indicators for death/exit and birth/entry dates. Our
function, \texttt{agents\_to\_aggregate()} does not require agents to
pass through every state listed and agents are allowed to start in any
state. This function is constrained to integer time steps (for example
days), but transformations (linear or otherwise) of the time columns can
account for finer grained time steps (e.g.~hours or minutes). The
function \texttt{agents\_to\_aggregate()} returns the total number of
individuals in each state at each time point. We view
\texttt{agents\_to\_aggregate()} as powerful tool in the step of
pre-processing data, one that quickly and simply transforms individual
level information into an aggregate view that can be processed and
analyzed in a simpler manner.

\textbf{EDA}

With raw data, exploratory analysis frequently means figuring out good
combinations of visualizations, numerical summaries, and groupings. An
expert coder can start from \texttt{agents\_to\_aggregate()} to
successfully accomplish EDA in many ways, but we have also developed
tools to allow a novice coder to rapidly explore data, as long as there
three unique epidemiological states (like the SIR model). Our
\texttt{geom\_aggregate()} provides a rapid way to explore different
subpopulations' experiences of the epidemic in the ternary lens. It
combines the ideas behind \texttt{agents\_to\_aggregate()} with
\pkg{ggplot2}'s \texttt{geom\_path()} and \pkg{ggtern}'s
\texttt{coord\_tern()} to visualize any number of groups epidemic
trajectory in 3d simplex \citep{Wickham2016, Hamilton2018}. We developed
these out-of-the-box visualization tools for SIR models because (1) SIR
models are some of the most common and basic epidemic state-based models
and (2) our simplex representation of these epidemics emphasizes a
``time-invarance'' representation of the data (see Section
\ref{sec:time-invariant}).

\textbf{Model Fitting and Simulations}

Although this package does not focus on estimating a model for the data,
we do provide some flexible functions for simulation of basic
discrete-time epidemic-state models with Bernoulli/Multinomial
transitions which can be used to (noisily) approximate ODEs which
describe transitions from one state to the next. These functions can be
naturally combined with \texttt{agents\_to\_aggregate()} to proceed in
the pipeline. The function \texttt{simulate\_SIR\_agents()} simulates an
SIR epidemic with user inputs for the number of simulations, the initial
number in each state, the infection and recovery parameters
\((\beta, \gamma\)), and the total number of discrete time steps. This
function allows for easy access to SIR model analysis and comparison.
Beyond SIR models, the function \texttt{simulate\_agents()} takes as
input a user-specified transition matrix and other epidemic parameters
to allow the user to create simulations of an outbreak for \textit{any}
number of states and any number of transitions among them. This allows
for users to explore the space of models in an intuitive way without
getting bogged down by too much mathematical detail. For consistency, we
have made output from \texttt{simulate\_agents()} and
\texttt{simulate\_SIR\_agents()} compatible with
\texttt{agents\_to\_aggregate()} so aggregate information may easily be
accessed.

\textbf{Post-processing}

Post-processing of modeling and simulation consists of making summary
statistics, plots, tables, and other ways to disseminate information to
the public. For example, comma separated value files (\texttt{.csv}) are
a standard way to share information within tables. However, model output
is often far more complicated than what a traditional \texttt{.csv}
would allow. As a result, a number of epidemic modeling packages return
a special class, specific to their modeling. The special classes often
contain a plethora of information from residuals, model diagnostics,
input parameters, and more. While incredibly useful, these special
classes can be difficult for novice coders to navigate.

To this end, we have adapted a series of \texttt{fortify}-style
functions, called \texttt{fortify\_aggregate()} which transform output
from packages like \pkg{pomp} and \pkg{EpiModel} into tidy-styled data
frames which contain information about the total number of individuals
in each state at a given time, for a given simulation. These fortify
functions have output that is consistent with that of
\texttt{agents\_to\_aggregate()}.

\textbf{Comparisons and Assessment}

Comparison and assessment of model fit or comparisons of one model to
another model can be performed in a variety of ways including mean
square error, AIC, plots, and more. Perhaps the most useful tool
\pkg{EpiCompare} has to offer to the expert, for comparison and
assessment of models, is in its post-processing tools which create a
standard output. It is then a matter of writing a script or function
made for that standard output to assess the results from multiple models
in the way the user desires.

However, for those who like more concrete tools, \pkg{EpiCompare} offers
functions to compare prediction regions to one another including
\texttt{geom\_prediction\_band()} (which plots the region), and
\texttt{create\_\{convex\_hull,delta\_ball\}\_structure()}(which returns
the \texttt{R} output for the given structure), and \texttt{contained()}
(which allows the user to determine if one set of points is contained in
a prediction band). Additionally, we offer ways to determine if model
outputs are compatible with one another, that is how extreme one output
is to another. \textcolor{orange}{Ben says something about distance}
